---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# dopingdata

<!-- badges: start -->

![](<https://img.shields.io/badge/lifecycle-experimental-orange.svg>)

<!-- https://img.shields.io/badge/lifecycle-stable-green.svg -->

<!-- badges: end -->

`dopingdata` contains data from the [United States Anti-Doping Agency](https://en.wikipedia.org/wiki/United_States_Anti-Doping_Agency) for exploration, modeling, and visualizations. The datasets in this package are derived from from the [USADA website](https://www.usada.org/) and the [World Anti-Doping Agency (WADA) banned substances list](https://www.wada-ama.org/en/prohibited-list?q=).

## Installation

You can install the development version of `dopingdata` like so:

``` r
# install.packages("devtools")
devtools::install_github("mjfrigaard/dopingdata")
```

## Raw Data

Each dataset was ha**rvest**ed using the [`rvest`](https://rvest.tidyverse.org/) and [`xml2`](https://xml2.r-lib.org/) packages, but using manners (with the [`polite`](https://dmi3kno.github.io/polite/) package).

-   All raw datasets have a date (`YYYY-MM-DD`) prefix and `_raw` suffix
    -   See the `scraping-usada-sanctions.Rmd` and `scraping-usada-proh-assoc.Rmd` vignettes for more information

    -   The raw data is then saved in the `inst/extdata/raw` folder

```{r raw-data, echo=FALSE}
fs::dir_tree("inst/extdata/raw")
```

## Processed Data

Processed datasets have the same dimensions and structure as the `_raw` data, but they've been formatted for easier wrangling/manipulation

-   All processed datasets have a `_pro` suffix

    -   Processed data have all had the column names formatted with `janitor::clean_names()`

    -   All of the text has been converted to lowercase

    -   See the `scraping-usada-sanctions.Rmd` and `scraping-usada-proh-assoc.Rmd` vignettes for more information

    -   Processed datasets are in the `inst/extdata/` folder:

```{r pro-data, echo=FALSE}
fs::dir_tree("inst/extdata/", regexp = "_pro")
```

## Derived Data

Additional datasets derived from the from the processed data and live in the `inst/extdata/` folder (and do not have a date prefix or any suffixes).

### Sanction Dates

The `sanction_announced` initially includes two dates (original, updated) for a selection of athletes. These observations are wrangled and formatted into dates. Derived from the `usada_sanctions_pro.csv` data.

-   See the `vignettes/sanction-dates.Rmd` vignette for more information:

    ```{r dates, echo = FALSE}
    fs::dir_tree("inst/extdata/", regexp = "_dates")
    ```

### Sports

The **`sport`** column has been 'tidied' and athletes/support personnel listed with more than one sport have been repeated in the data. These data are derived from the `usada_dates.csv` data.

-   the sports separated and 'tidied' in `vignettes/sanction-sports.Rmd`:

    ```{r sports, echo = FALSE}
    fs::dir_tree("inst/extdata/", regexp = "sports")
    ```

### Adverse Analytical Findings (AAF)

The **Single Substance Violations** (i.e., Adverse Analytical Finding, AAF) are derived from `tidy_sports.csv`:

-   Adverse analytical finding (AAF) sanctions with a single substance are in `vignettes/single-aaf-substances.Rmd`:

    ```{r single_aafs, echo = FALSE}
    fs::dir_tree("inst/extdata/", regexp = "single_aafs")
    ```

The **Multiple Substance Violations** (i.e., Adverse Analytical Finding, AAF) are derived from

-   Adverse analytical finding (AAF) sanctions with multiple substance are in `vignettes/multiple-aaf-substances.Rmd`:

    ```{r multiple_aaf, echo=FALSE}
    fs::dir_tree("inst/extdata/", regexp = "tidy_mult")
    ```

### Non-Analytical Anti-doping Rule Violation (ADRV)

The **Single Substance, Non-Analytic Violations** (i.e., Non-Analytical Anti-doping Rule Violation, ADRV) are derived from the `tidy_sports.csv` data.

The **Multiple Substance, Non-Analytic Violations** (i.e., Non-Analytical Anti-doping Rule Violation, ADRV) are derived from the `tidy_sports.csv` data.

### Sanction terms

The type and length of suspension are derived `tidy_single_aafs.csv` and `tidy_multp_aaf_subs.csv` data.

### Prohibited Associations

The prohibited associations are derived from the `usada_prohib_assoc_pro.csv` data.

## WADA data

This package also includes a list of the [World Anti-Doping Agency (WADA)](https://en.wikipedia.org/wiki/World_Anti-Doping_Agency) banned substances (published online [here](https://www.wada-ama.org/en/prohibited-list?q=)).

```{r wada-data, echo=FALSE}
fs::dir_tree("inst/extdata/wada/")
```

### More WADA Resources on banned substances

-   [USADA - UFC](https://ufc.usada.org/wp-content/uploads/UFC-Prohibited-List.pdf)
