---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# `dopingdata`

<!-- badges: start -->
<!-- badges: end -->

The goal of `dopingdata` is to provide datasets from the [United States Anti-Doping Agency](https://en.wikipedia.org/wiki/United_States_Anti-Doping_Agency) for exploration, modeling, and visualizations. 

The datasets in this package are derived from from the [USADA website](https://www.usada.org/) and the [World Anti-Doping Agency (WADA) banned substances list](https://www.wada-ama.org/en/prohibited-list?q=). 


## Installation

You can install the development version of `dopingdata` like so:

``` r
# install.packages("devtools")
devtools::install_github("mjfrigaard/dopingdata")
```

## Harvesting Data 

Each dataset was ha**rvest**ed using the [`rvest`](https://rvest.tidyverse.org/) and [`xml2`](https://xml2.r-lib.org/) packages, but using manners (with the [`polite`](https://dmi3kno.github.io/polite/) package). See the `Scraping USADA sanctions` and `Scraping USADA prohibited associations` vignettes for more information. 

The raw data is then saved in the `inst/extdata/raw` folder. Raw datasets have a date (`YYYY-MM-DD`) prefix and `_raw` suffix.

```{r raw-data, echo=FALSE}
fs::dir_tree("inst/extdata/raw")
```


## Processed Data 

The processed datasets are created from the scraped raw data and are structured for different purposes. In all cases, the vignettes contain the steps and custom functions used to harvest each dataset. All processed datasets have a `_pro` suffix. 

Processed datasets have the same dimensions and structure as the `_raw` data, but they've been formatted for easier wrangling/manipulation. 

Processed datasets are in the `inst/extdata/` folder:

```{r pro-data, echo=FALSE}
fs::dir_tree("inst/extdata/", regexp = "_pro")
```

## Derived Data 

The following datasets are derived from the processed data: 

1. Sanction Dates  
2. Sports  
3. Substances (AAF)  
4. Non-Analytic Sanctions (ADRV)   
5. Sanction Terms  
6. Prohibited Associations  
7. Athletes  

All derived datasets are in the `inst/extdata/` folder (and do not have a date prefix or any suffixes).

### Sanction Dates  

Derived from `usada_sanctions_pro.csv`: 

  - the dates are wrangled and formatted in `vignettes/sanction-dates.Rmd`: 
   
    ```{r dates, echo = FALSE}
    fs::dir_tree("inst/extdata/", regexp = "_dates")
    ```

### Sports 

Derived from `usada_dates.csv`:

  - the sports separated and 'tidied' in `vignettes/sanction-sports.Rmd`: 
   
    ```{r sports, echo = FALSE}
    fs::dir_tree("inst/extdata/", regexp = "sports")
    ```

### Substances (AAF)

Derived from `tidy_sports.csv`:

  - Adverse analytical findings with a single substance are in `vignettes/sanction-aaf-substances.Rmd`: 
   
    ```{r aaf, echo = FALSE}
    fs::dir_tree("inst/extdata/", regexp = "aaf")
    ```

### Non-Analytic Sanctions (ADRV)

### Sanction terms

### Prohibited Associations  

### Athletes  

## WADA data

This package also includes a list of the [World Anti-Doping Agency (WADA)](https://en.wikipedia.org/wiki/World_Anti-Doping_Agency) banned substances (published online [here](https://www.wada-ama.org/en/prohibited-list?q=)).

```{r wada-data, echo=FALSE}
fs::dir_tree("inst/extdata/wada/")
```

### More WADA Resources on banned substances 

- [USADA - UFC](https://ufc.usada.org/wp-content/uploads/UFC-Prohibited-List.pdf)



