---
output: 
  rmarkdown::github_document:
    df_print: tibble
    toc: true
    toc_depth: 4
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
options(
  repos = "https://cloud.r-project.org")
```

<!-- badges: start -->

![](https://img.shields.io/badge/lifecycle-experimental-orange.svg)

<!-- https://img.shields.io/badge/lifecycle-stable-green.svg -->

<!-- badges: end -->

`dopingdata` contains data from the [United States Anti-Doping Agency](https://en.wikipedia.org/wiki/United_States_Anti-Doping_Agency) for exploration, modeling, and visualizations. The datasets in this package are derived from from the [USADA website](https://www.usada.org/) and the [World Anti-Doping Agency (WADA) banned substances list](https://www.wada-ama.org/en/prohibited-list?q=). Scraping, processing, and visualizing these data presented so many unique challenges I decided to combine the utilities into a package.

## Installation

You can install the development version of `dopingdata` like so:

``` r
# install.packages("devtools")
devtools::install_github("mjfrigaard/dopingdata")
```

```{r}
library(dopingdata)
```

## Scraping Sanction Data

Each dataset was ha**rvest**ed using the [`rvest`](https://rvest.tidyverse.org/) and [`xml2`](https://xml2.r-lib.org/) packages, but using manners (with the [`polite`](https://dmi3kno.github.io/polite/) package).

```{r , eval=FALSE}
scrape_sanctions(dest_path = "dev")
```

```         
Exporting raw data: dev/2023-12-19/2023-12-19-usada_sanctions_raw.csv
Raw data successfully exported!
```

## Processing Data

`process_text()` performs the following:

-   Processed data have all had the column names formatted `snake_case`

-   All of the text has been converted to lowercase

```{r}
usada_raw <- read.csv("dev/2023-12-19/2023-12-19-usada_sanctions_raw.csv")
str(usada_raw)
```

```{r}
usada <- process_text(raw_data = usada_raw)
str(usada)
```



## Sanction Dates

`sanction_announced` contains the date the sanction was announced, and about 30 of these contain two values (`original` and `updated`). Wrangling these values pose some challenges because they aren't *consistently* messy. 

```{r}
subset(usada, 
  grepl("^original", usada[['sanction_announced']]), 
  sanction_announced)
```

I decided to split these into three `data.frame`s to wrangle the dates (`bad_dates`, `good_dates`, and `no_dates`).

```{r}
bad_dates <- subset(usada, 
  grepl("^original", usada[['sanction_announced']]))
good_dates <- subset(usada, 
  !grepl("^original", usada[['sanction_announced']]) & sanction_announced != "")
no_dates <- subset(usada,
  athlete == "*name removed" & sanction_announced == "")
```

I've written a `clean_dates()` function that takes `date_col`, `split` and `pattern` arguments: 

+ `df` = processed USADA dataset with messy dates

+ `date_col` = sanction date column (usually `sanction_announced`)

+ `split` = regex to pass to split argument of `strsplit()` (defaults to `"updated"`)

+ `pattern` = regex for other non-date pattern (defaults to `"original"`)

```{r}
cleaned_dates <- clean_dates(
  df = bad_dates, 
  date_col = "sanction_announced", 
  split = "updated", 
  pattern = "original")
head(
  cleaned_dates[c('sanction_announced', 'pattern_date', 'split_date')]
)
```

I split the bad dates on `"updated"` and provided `"original"` as the pattern (the opposite will also work). The `sanction_date` column will contain the correctly formatted updated `sanction_date`.  

```{r}
names(cleaned_dates)[names(cleaned_dates) == 'split_date'] <- 'sanction_date'
names(cleaned_dates)[names(cleaned_dates) == 'pattern_date'] <- 'original_date'
head(
  cleaned_dates[c('sanction_announced', 'sanction_date', 'original_date')]
)
```

After formatting `good_dates` and removing `original_date` column we can combine the two with `rbind()`:

```{r}
good_dates$sanction_date <- as.Date(x = good_dates[['sanction_announced']], 
                                    format = "%m/%d/%Y")
# get intersecting names 
nms <- intersect(names(cleaned_dates), names(good_dates))
# bind the two datasets 
usada_dates <- rbind(good_dates, cleaned_dates[nms])
str(usada_dates)
```


```{r}
export_data(x = usada_dates, path = "data-raw")
```

## Sports 

To wrangle the sports, I'll use packages and functions from the `tidyverse` (`dplyr`, `stringr`, `tidyr`, etc.), but I also provide the base R alternatives (wherever possible).

```{r}
library(tidyverse)
```

`tidyverse` functions will return a `tibble` (not a `data.frame`), which prints fewer rows to the console. 

```{r}
usada_sports <- tibble::as_tibble(usada_dates)
str(usada_sports)
```

We can start by counting the `sport` column:

```{r count-sports-raw}
usada_sports |> 
  dplyr::count(sport, sort = TRUE)
```

### Support personnel

Some of the sports aren't sports--they're `athlete support personnel`. These have `support_personnel` identifier.

```{r support_personnel}
usada_sports <- dplyr::mutate(usada_dates, 
  # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(
      sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA)) 

usada_sports |> 
  dplyr::count(sport, support_personnel) |> 
  tidyr::pivot_wider(names_from = support_personnel, values_from = n)
```

### `track and field` or  `track & field`

I'm going to convert sports like `track & field` as `track and field` (this will help me  determine which athletes/support personnel are involved in multiple sports). 

```{r track-field}
usada_sports <- dplyr::mutate(usada_sports,
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field'))

usada_sports |> 
  dplyr::count(sport, support_personnel) |> 
  tidyr::pivot_wider(names_from = support_personnel, values_from = n)
```


### Spelling 

The incorrect spelling for `brazilian jiu-jitsu` (`brazillian jiu-jitsu`) is corrected below.

```{r bjj}
usada_sports <- dplyr::mutate(usada_sports, 
  # brazilian jiu-jitsu
  sport = dplyr::case_when(
    sport == 'brazillian jiu-jitsu' ~ 'brazilian jiu-jitsu',
    TRUE ~ sport)) 
usada_sports |> 
  dplyr::filter(stringr::str_detect(sport, "jitsu")) |> 
  dplyr::count(sport, sort = TRUE)
```


### `paralympic`

An identifier for paralympic sports: `paralympic`. 

```{r paralympic}
usada_sports <- dplyr::mutate(usada_sports, 
  # paralympic
  paralympic = 
    dplyr::if_else(condition = stringr::str_detect(sport, "paralympic|para"), 
      true = TRUE, false = FALSE, missing = NA)) 

usada_sports |> 
  dplyr::count(paralympic, sport) |> 
  tidyr::pivot_wider(names_from = paralympic, values_from = n)
```

### `multiple_sports` 

Now we can identify the multiple sports using `and` and `, ` in the regular expression. 

```{r multiple_sports}
usada_sports <- dplyr::mutate(usada_sports, 
  # multiple_sports
  multiple_sports = 
    if_else(condition = stringr::str_detect(sport, "and |, "), 
      true = TRUE, false = FALSE, missing = NA))

usada_sports |> 
  dplyr::count(multiple_sports, sport) |> 
  tidyr::pivot_wider(names_from = multiple_sports, values_from = n)
```

### Tidy sports

The athletes listed with multiple sports will occupy multiple rows in a 'tidy' version of `usada_sports`. We can check this using `count()` below to verify we have one athlete for each sport:

```{r check-tidy-sports}
usada_sports |> 
  dplyr::filter(multiple_sports == TRUE) |> 
  tidyr::separate_rows(sport, sep = "and|, ") |> 
  dplyr::count(athlete, sport)
```

Remove any white-space left over from `tidyr::separate_rows()` with `stringr::str_trim()` to create a `multp_usada_sports`. 

```{r multp_usada_sports}
multp_usada_sports <- usada_sports |> 
  dplyr::filter(multiple_sports == TRUE) |> 
  tidyr::separate_rows(sport, sep = "and|, ") |> 
  dplyr::mutate(sport = stringr::str_trim(sport, side = "both"))
glimpse(multp_usada_sports)
```

To combine these with the single-sport athletes, I'll need to remove the single occurrences of the multiple-sport athletes in `usada_sports` (and replace them with the `tidy_usada_sports`)

```{r single_usada_sports}
# get athlete names
multp_sports_athletes <- unique(multp_usada_sports['athlete'])
# remove from 
single_usada_sports <- dplyr::filter(usada_sports, 
                                        athlete %nin% multp_sports_athletes)
```

Verify there aren't any duplicates in `single_usada_sports` (there should be zero rows here)

```{r check-single_usada_sports}
single_usada_sports |> 
  dplyr::count(athlete, sanction_date, sport) |> 
  dplyr::filter(n > 1)
```

Now we can combine the two datasets and verify there aren't any duplicates (again). 

```{r tidy_sports}
tidy_sports <- dplyr::bind_rows(single_usada_sports, multp_usada_sports)
tidy_sports |> 
  dplyr::count(athlete, sanction_date, sport) |> 
  dplyr::filter(n > 1)
```

```{r check-tidy_sports}
str(tidy_sports)
```

Export `tidy_sports` to `data-raw/`

```{r export-sports, eval=TRUE}
export_data(
  x = tidy_sports, 
  path = "data-raw")
```

Verify 

```{r}
fs::dir_tree("data-raw/2023-12-19")
```

## Substances 

We'll start by classifying 'adverse analytic findings' for a single banned substance. I've written a `get_recent_file()` function to quickly import .csv files from a specified directory:

```{r}
get_recent_file("data-raw/2023-12-19/", regex = 'sports', ext = '.csv')
```

```{r}
tidy_sports <- read.delim(file = 'data-raw/2023-12-19/2023-12-19-tidy_sports.csv', sep = ',')
```

#### AAFs vs ADRVs

The sanctions are divided into two categories: 

- `analytic`: [Adverse Analytical Finding](https://www.usada.org/spirit-of-sport/education/alphabet-soup-results-management/#:~:text), AAF; *An AAF is a report from a WADA-accredited laboratory that identifies the presence of a prohibited substance and/or its metabolites or markers in a sample.* 

- `non-analytic`: [Non-Analytical Anti-doping Rule Violation](https://www.usada.org/spirit-of-sport/education/non-analytical-anti-doping-rule-violations/) ADRV; *a non-analytical anti-doping rule violation does not stem from a positive urine or blood sample, but instead originates from, and is substantiated by, other evidence of doping or violations by an athlete or athlete support personnel.*. 

##### `substance_reason`

The `substance_reason` column contains the details of the sanction, which can include the following information: 

1. The name of the banned substance  

2. A description of the infraction (if non-analytic)  

We will use regular expressions to identify the type of substance behind the sanction. See the examples below:

```{r steroid-view}
stringr::str_view(tidy_sports$substance_reason, 
  "use \\(epo & hgh\\)", match = TRUE)
```

```{r non-analytic-view}
stringr::str_view(tidy_sports$substance_reason,
  "tampering, complicity", match = TRUE)
```


### Sanction type

Most of the non-analytic sanctions include the terms `non-analytic`/`non-analytical`/etc., as a prefix in the `substance_reason` column.

I can pass these terms as regular expressions to create the `sanction_type` variable, which will contain two values: `non-analytical` and `analytical`. I'll save this variable in a new the `usada_substances` dataset:

```{r usada_sanction_types, eval=TRUE}
usada_substances <- dplyr::mutate(.data = tidy_sports,
    sanction_type = dplyr::case_when(
      stringr::str_detect(string = substance_reason,
        "non-analytical") ~ "non-analytical",
      !stringr::str_detect(substance_reason,
        "non-analytical") ~ "analytical",
      TRUE ~ NA_character_
    )
  )
usada_substances |>
  dplyr::count(sanction_type, sort = TRUE)
```


Now I can filter `usada_substances` to the `analytical` sanctions in `sanction_type`.

### Multiple vs. single substances

*How can I identify the single vs. multiple substances?*

Let's take a look at four different sanctions in `sanction_type_example`:

```{r sanction_type_example, echo=FALSE}
sanction_type_example <- tibble::tibble(
             sport = c("swimming","track & field",
                       "triathlon","track & field"),
  substance_reason = c("non-analytical: 3 whereabouts failures","cannabinoids",
                       "androgenic anabolic steroid; cannabinoids; anastrozole; clomiphene; gw1516",
                       "non-analytical: tampering, administration (iv), and trafficking (testosterone)"),
     sanction_type = c("non-analytical","analytical",
                       "analytical","non-analytical"))
sanction_type_example
```

We can see two analytic and two non-analytic sanctions, and each one has a single or multiple substance/reason. Fortunately, the sanctions with multiple items are separated by either semicolons (`; `), commas (`, `), or a conjunction (`and`), so I can use a regular expression to separate the items.

```{r test-substance_cat}
dplyr::mutate(sanction_type_example,
  substance_cat = case_when(
    # identify the multiple_sr substances using a regular expression
    stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'multiple',
    # negate the regular expression for the single substances
    !stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'single',
    TRUE ~ NA_character_)) |>
  dplyr::count(substance_cat, sanction_type) |> 
  tidyr::pivot_wider(names_from = substance_cat, values_from = n)
```

The `substance_cat` identifier can be used to separate sanctions with multiple substance/reasons from sanctions with a single substance or reason.

```{r}
usada_substances <- usada_substances |>
  dplyr::mutate(substance_cat = dplyr::case_when(
    stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'multiple',
    !stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'single',
    TRUE ~ NA_character_))
usada_substances |> 
  dplyr::count(substance_cat, sort = TRUE)
```

## Single analytic substances

First we'll focus on the sanctions with a single substance listed (we will deal with multiple substances next).

```{r single_analytic_substances}
single_analytic_substances <- usada_substances |>
  dplyr::filter(substance_cat == 'single' & sanction_type == "analytical")
head(single_analytic_substances)
```
## Classifying substances with `classify_wada_substances()`

To identify the [WADA banned substances](https://www.wada-ama.org/en/prohibited-list#search-anchor), I've written `classify_wada_substances()`, a function that scans the `substance_reason` column and identifies any substances found on the [WADA list](https://www.wada-ama.org/sites/default/files/2022-09/2023list_en_final_9_september_2022.pdf). See the `classify_wada_substances()` documentation for more information.

### How `classify_wada_substances()` works (S1 ANABOLIC AGENTS example)

`classify_wada_substances()` creates a `substance_group` variable with each of the WADA classifications (these are stored in `dopingdata::wada_classes`)

```{r show-dopingdata-wada_classes}
dopingdata::wada_classes
```


`dopingdata` stores vectors with each substance group in the WADA list (S1 substances are below):

```{r s1_substances}
head(dopingdata::s1_substances, 10)
```

The substance group vectors are also stored as regular expressions (`s1_regex`), which we can use to match the `substance_reason` column on:

```{r veiw-s1}
stringr::str_view(string = single_analytic_substances$substance_reason,
  pattern = s1_regex, match = TRUE)
```

The output from `classify_wada_substances()` can be used to answer questions like: *what `substance_group`'s appear the most?*

```{r demo_classify_wada_substances}
single_analytic_substances <- classify_wada_substances(
  usada_data = single_analytic_substances,
  subs_column = substance_reason) 
single_analytic_substances |>
  dplyr::count(substance_group, sort = TRUE)
```

## Unclassified substances

The substances that were not classified using the standard WADA list can be added with `classify_substance()`

```{r unclass_single_aaf_substances}
single_analytic_substances |>
  filter(substance_group == "UNCLASSIFIED") |>
  count(substance_reason, sort = TRUE)
```

### Re-classify substances 

`classify_substance()` takes a `df`, `substance`, and `value`:

```{r}
single_analytic_substances <- classify_substance(
  df = single_analytic_substances,
  substance = "elevated t/e",
  value = "S1 ANABOLIC AGENTS")
single_analytic_substances |>
  dplyr::filter(substance_reason == "elevated t/e") |>
  dplyr::select(substance_reason, substance_cat, substance_group)
```

