---
title: "Single ADRV sanctions"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Single ADRV sanctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(
  pillar.print_max = 25, 
  dplyr.print_max = 25,
  knitr.kable.NA = '',
  pillar.print_min = 25)
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

Packages:

```{r pkg, message=FALSE, warning=FALSE}
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

## Motivation 

This vignette covers the non-analytic, single substance violations in the [USADA sanction table](https://www.usada.org/testing/results/sanctions/). 

## Data versioning 

`dopingdata` uses data from tables that get updated regularly. To keep track of the versions, I've separated the raw (`_raw`) and processed (`_pro`) data in the `inst/` folder: 

```{r}
fs::dir_tree("../inst/extdata", recurse = FALSE)
```


I also have a helper function for getting the recent version:

```{r tdy_stmp}
get_recent("../inst/extdata/", full = TRUE)
get_recent("../inst/extdata/raw", full = TRUE)
```


## Data 

Below I'll import the most recent version of the tidy USADA sports data:

```{r get_recent_path-tidy-sports, eval=FALSE}
get_recent_path("../inst/extdata/", 
  type = "csv")
✔ import code pasted to clipboard!
✔ use: readr::read_csv('../inst/extdata/tidy_sports.csv')
```

```{r import-tidy_usada_sports}
tidy_usada_sports <- readr::read_csv(
  file = '../inst/extdata/tidy_sports.csv')
dplyr::glimpse(tidy_usada_sports)
```

These data have had their names standardized, and all the text has been converted to lowercase. 

## AAFs vs ADRVs

The sanctions are divided into two categories: 

- `analytic`: [Adverse Analytical Finding](https://www.usada.org/spirit-of-sport/education/alphabet-soup-results-management/#:~:text), AAF; *An AAF is a report from a WADA-accredited laboratory that identifies the presence of a prohibited substance and/or its metabolites or markers in a sample.* 

- `non-analytic`: [Non-Analytical Anti-doping Rule Violation](https://www.usada.org/spirit-of-sport/education/non-analytical-anti-doping-rule-violations/) ADRV; *a non-analytical anti-doping rule violation does not stem from a positive urine or blood sample, but instead originates from, and is substantiated by, other evidence of doping or violations by an athlete or athlete support personnel.*. 
