---
title: "Classifying substances"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Classifying substances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r pkgs, eval=FALSE}
pkgs <- c("dplyr", "stringr", "tidyr", "forcats")
install.packages(pkgs, quiet = TRUE)
```


```{r setup, message=FALSE, warning=FALSE}
library(dopingdata)
library(dplyr)
library(stringr)
library(tidyr)
library(forcats)
```

This vignette covers classifying 'adverse analytic findings' for a single banned substance. I've written a `get_recent_file()` function to quickly import `.csv` files from a specified directory:

```{r}
pth <- system.file("extdata", "demo", package = "dopingdata")
get_recent_file(pth, regex = 'sports', ext = '.csv')
```

This makes it easy to paste the necessary import code into the console (or R markdown file):

```{r}
tidy_sports <- read.delim(file = '/Library/Frameworks/R.framework/Versions/4.3-x86_64/Resources/library/dopingdata/extdata/demo/2023-12-21-tidy_sports.csv', sep = ',')
```

## AAFs vs ADRVs

The sanctions are divided into two categories: 

- `analytic`: [Adverse Analytical Finding](https://www.usada.org/spirit-of-sport/education/alphabet-soup-results-management/#:~:text), AAF; *An AAF is a report from a WADA-accredited laboratory that identifies the presence of a prohibited substance and/or its metabolites or markers in a sample.* 

- `non-analytic`: [Non-Analytical Anti-doping Rule Violation](https://www.usada.org/spirit-of-sport/education/non-analytical-anti-doping-rule-violations/) ADRV; *a non-analytical anti-doping rule violation does not stem from a positive urine or blood sample, but instead originates from, and is substantiated by, other evidence of doping or violations by an athlete or athlete support personnel.*. 

### Substance/reason

The `substance_reason` column contains the details of the sanction, which can include the following information: 

1. The name of the banned substance  

2. A description of the infraction (if non-analytic)  

We will use regular expressions to identify the type of substance behind the sanction. See the examples below:

```{r steroid-view}
stringr::str_view(tidy_sports[['substance_reason']], 
  "use \\(epo & hgh\\)", match = TRUE)
```

```{r non-analytic-view}
stringr::str_view(tidy_sports[['substance_reason']],
  "tampering, complicity", match = TRUE)
```

Most of the non-analytic sanctions include the terms `non-analytic`/`non-analytical`/etc., as a prefix in the `substance_reason` column.

### Sanction types

I can pass these terms as regular expressions to create the `sanction_type` variable, which will contain two values: `non-analytic` and `analytic`. I'll save this variable in a new intermediate `substances` dataset:

```{r usada_sanction_types, eval=TRUE}
substances <- dplyr::mutate(.data = tidy_sports,
    sanction_type = dplyr::case_when(
      stringr::str_detect(string = substance_reason,
        "non-analytical") ~ "non-analytic",
      !stringr::str_detect(substance_reason,
        "non-analytical") ~ "analytic",
      TRUE ~ NA_character_
    )
  )
substances |>
  dplyr::count(sanction_type, sort = TRUE)
```


Now I can filter `substances` to the `analytical` sanctions in `sanction_type`.

#### *How can I identify the single vs. multiple substances?*

Let's take a look at four different sanctions in `example_sanction_type`:

```{r example_sanction_type, echo=FALSE}
dopingdata::example_sanction_type
```

### Substance category

We can see two analytic and two non-analytic sanctions, and each one has a single and multiple substance/reason. Fortunately, the sanctions with multiple items are separated by either semicolons (`; `), commas (`, `), or a conjunction (`and`), and can be separated by a regular expression:

```{r test-substance_cat}
dplyr::mutate(example_sanction_type,
  substance_cat = dplyr::case_when(
    # identify the multiple_sr substances using a regular expression
    stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'multiple',
    # negate the regular expression for the single substances
    !stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'single',
    TRUE ~ NA_character_)) |>
  dplyr::count(substance_cat, substance_reason) |> 
  tidyr::pivot_wider(names_from = substance_cat, values_from = n)
```

The `substance_cat` identifier can be used to separate sanctions with multiple substance/reasons from sanctions with a single substance or reason.

```{r}
substances <- substances |>
  dplyr::mutate(substance_cat = dplyr::case_when(
    stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'multiple',
    !stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ 'single',
    TRUE ~ NA_character_))
substances |> 
  dplyr::count(substance_cat, sort = TRUE)
```

## Single analytic substances

First create a dataset that contains the sanctions with a *single* substance listed. Store these in `single_analytic_substances`.

```{r single_analytic_substances}
single_analytic_substances <- substances |>
  dplyr::filter(substance_cat == 'single' & sanction_type == "analytic")
```

View the top ten single analytic substances:

```{r}
single_analytic_substances |> 
  dplyr::count(substance_reason, sort = TRUE) |> 
  head(10)
```

## Multiple analytic substances

Next create a dataset with the sanctions listing *multiple* substances in `substance_reason`. Store these in `multiple_analytic_substances`.

```{r multiple_analytic_substances}
multiple_analytic_substances <- substances |>
  dplyr::filter(substance_cat == 'multiple' & sanction_type == "analytic")
```

View the top ten multiple analytic substances:

```{r}
multiple_analytic_substances |> 
  dplyr::count(substance_reason, sort = TRUE) |> 
  head(10)
```

### Tidying substances 

Tidying the sanctions with multiple [WADA banned substances](https://www.wada-ama.org/en/prohibited-list?q) (i.e., one substance per athlete per row) will result in certain athletes appearing in the dataset more than once. The regular expressions below cover a range of semicolons, tabs, and spaces to identify and separate each substance. 

#### add_match_col()

> *I've written the `add_match_col()` function, which creates a new `'matched'` column with the matched regular expression pattern (it's like `stringr::str_view()`, but in a `data.frame`/`tibble`). I used `add_match_col()` while determining the correct pattern to match on (i.e., any substances listing `metabolites`):*

```{r}
dplyr::mutate(example_sanction_type,
  # add matched column
  punct_match = add_match_col(
    string = substance_reason, 
    pattern = "[[:punct:]]")) |> 
  dplyr::select(substance_reason, dplyr::last_col()) 
```

> *The rows above are all matching correctly on the regular expression pattern.* 

Below are a series of regular expression to 1) match the substances that list `metabolites` 2) differentiate the multiple substances, and 3) trims the white space from the tidy `substance_reason` column.

The code below tests this pattern on a sample from `multiple_analytic_substances` before its passed to `tidyr::separate_rows()`:

```{r}
dplyr::sample_n(multiple_analytic_substances, size = 10, replace = FALSE) |> 
  dplyr::mutate(
    # replace plurals
      substance_reason = stringr::str_replace_all(substance_reason,
        "and its metabolite|and its metabolites|its metabolite",
        "(metabolite)")) |> 
    tidyr::separate_rows(substance_reason, 
        sep = "; |;\t|\\t|, |;| and |and a |and | & | / ") |> 
    dplyr::mutate(substance_reason = trimws(substance_reason, "both")) |> 
    dplyr::select(athlete, substance_reason)
```

After confirming the pattern is working, the output will be stored in `tidy_multiple_substances`.

```{r tidy_multiple_substances}
tidy_multiple_substances <- dplyr::mutate(multiple_analytic_substances,
  # replace plurals
    substance_reason = stringr::str_replace_all(substance_reason,
      "and its metabolite|and its metabolites|its metabolite",
      "(metabolite)")) |> 
  tidyr::separate_rows(substance_reason, 
      sep = "; |;\t|\\t|, |;| and |and a |and | & | / ") |> 
  dplyr::mutate(substance_reason = trimws(substance_reason, "both"))
```

With both single and multiple substances in tidy format, they can be combined together into a single `tidy_substances` dataset.

```{r tidy_substances}
tidy_substances <- rbind(single_analytic_substances, tidy_multiple_substances) 
```

The top 10 tidy substances are below: 

```{r}
tidy_substances |> 
  dplyr::count(substance_reason, sort = TRUE) |> 
  head(10)
```

```{r, include=FALSE, eval=FALSE}
dplyr::mutate(multiple_analytic_substances,
  # replace plurals
    substance_reason = stringr::str_replace_all(substance_reason,
      "and its metabolite|and its metabolites|its metabolite",
      "(metabolite)")) |> 
  tidyr::separate_rows(substance_reason, 
      sep = "; |;\t|\\t|, |;| and |and a |and | & | / ") |> 
  dplyr::mutate(substance_reason = trimws(substance_reason, "both")) |> 
  dplyr::select(substance_reason) |> 
  purrr::as_vector() |> 
  writeLines()
```

## Classifying substances 

To identify the [WADA banned substances](https://www.wada-ama.org/en/prohibited-list#search-anchor), I've written `classify_wada_substances()`, a function that scans the `substance_reason` column and identifies any substances found on the [WADA list](https://www.wada-ama.org/sites/default/files/2022-09/2023list_en_final_9_september_2022.pdf). See the `classify_wada_substances()` documentation for more information.

### WADA Classes 

`classify_wada_substances()` creates a `substance_group` variable with each of the WADA classifications (stored in `dopingdata::wada_classes`):

```{r show-dopingdata-wada_classes}
dopingdata::wada_classes
```


`dopingdata` stores vectors with each substance group in the WADA list (the `S1 ANABOLIC AGENTS` substances are below):

```{r s1_substances}
head(dopingdata::s1_substances, 10)
```

The substance group vectors are passed to make_regex() to create a regular expressions (`s1_regex`), which we can use to match the `substance_reason` column on (see the example using `dopingdata::example_tidy_substances` dataset):

```{r veiw-s1}
s1_regex <- make_regex(x = dopingdata::s1_substances)
stringr::str_view(string = example_tidy_substances$substance_reason,
  pattern = s1_regex, match = TRUE)
```

The output from `classify_wada_substances()` can be used to answer questions like: *what `substance_group`'s appear the most?*

```{r demo_classify_wada_substances}
tidy_substances <- classify_wada_substances(
  usada_data = tidy_substances,
  subs_column = "substance_reason") 
```

```{r REMOVE-SUBSTANCE-DATA, echo=FALSE}
remove(list = ls()[grep(pattern = "mult|single|pth|regex", ls())])
```


### **UNCLASSIFIED** single substances 

The following `single` substances are marked as `UNCLASSIFIED`:

```{r}
tidy_substances |>
  dplyr::filter(
      substance_cat == "single" & 
      substance_group == "UNCLASSIFIED" & 
      substance_reason != "") |>
  dplyr::distinct(athlete, substance_reason)
```

The final unclassified substance is actually a result from a miss-classified sanction type (for `rodriguez, yair`). 

```{r}
tidy_substances |>
  dplyr::filter(athlete == "rodriguez, yair") |>
  dplyr::select(athlete, substance_reason, substance_group, sanction_type)
```

For this particular athlete, 1) the `sanction_type` should be `non-analytic`, and 2) the `substance_group` should be missing (`NA_character_`)

```{r}
tidy_substances <- tidy_substances |>
  dplyr::mutate(sanction_type = dplyr::case_when(
    athlete == "rodriguez, yair" ~ "non-analytic",
    TRUE ~ sanction_type
  )) |> 
  dplyr::mutate(substance_group = dplyr::case_when(
    athlete == "rodriguez, yair" ~ NA_character_,
    TRUE ~ substance_group
  )) 
tidy_substances |>
  dplyr::filter(athlete == "rodriguez, yair") |>
  dplyr::select(athlete, substance_reason, substance_group, sanction_type)
```

### **UNCLASSIFIED** multiple substances 

```{r}
tidy_substances |>
  dplyr::filter(
      substance_cat == "multiple" & 
      substance_group == "UNCLASSIFIED" & 
      substance_reason != "") |>
  dplyr::distinct(substance_reason)
```

## Re-classifying substances

Any substances that are not classified from the existing WADA list can be added with `reclass_substance()` (these substances can also added to their relative vector and regular expression in `data-raw/`)

### **reclass_substance()**

`reclass_substance()` takes a `df`, `substance`, and `value`:

```{r , eval=FALSE, echo=FALSE}
tidy_substances |>
  dplyr::filter(
      substance_cat == "multiple" & 
      substance_group == "UNCLASSIFIED" & 
      substance_reason != "") |>
  dplyr::distinct(substance_reason) |> 
  purrr::as_vector() |> 
  writeLines()
```

```{r}
# 2a-methyl-5a-androstan-3a-ol-17-one ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=arimistane&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "^2a-methyl-5a-androstan-3a-ol-17-one$",
  value = "S1 ANABOLIC AGENTS")
# d-methamphetamine ----
# https://www.usada.org/sanction/hillary-tran-accepts-doping-sanction/
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "d-methamphetamine",
  value = "S6 STIMULANTS")
# arimistane ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=arimistane&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "arimistane",
  value = "S4 HORMONE AND METABOLIC MODULATORS")
# torasemide ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=torasemide&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "torasemide",
  value = "S5 DIURETICS/MASKING AGENTS")
# igf-1 ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=igf-1&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "igf-1",
  value = "S2 PEP HORMONES/G FACTORS/MIMETICS")
# human chorionic gonadotropin (hcg)  ----
# https://www.usada.org/athletes/antidoping101/athlete-guide-anti-doping/
# https://www.usada.org/spirit-of-sport/education/wellness-and-anti-aging-clinics/
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "^human chorionic gonadotropin \\(hcg\\)$",
  value = "S2 PEP HORMONES/G FACTORS/MIMETICS")
# intact human chorionic gonadtrophin (hcg)  ----
# https://www.usada.org/athletes/antidoping101/athlete-guide-anti-doping/
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "^intact human chorionic gonadtrophin \\(hcg\\)$",
  value = "S2 PEP HORMONES/G FACTORS/MIMETICS")
# human chorionic gonadotrophin (hcg)  ----
# https://www.usada.org/athletes/antidoping101/athlete-guide-anti-doping/
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "human chorionic gonadotrophin \\(hcg\\)",
  value = "S2 PEP HORMONES/G FACTORS/MIMETICS")
# aod-9064  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=aod-9064&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "aod-9064",
  value = "S2 PEP HORMONES/G FACTORS/MIMETICS")
# s-23  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=s-23&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "s-23",
  value = "S1 ANABOLIC AGENTS")
# methenolone  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=methenolone&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "methenolone",
  value = "S1 ANABOLIC AGENTS")
# thiazide metabolite 4-amino-6-chloro-1,3-benzenedisulfonamide (acb)  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=thiazide&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "^thiazide metabolite 4-amino-6-chloro-1,3-benzenedisulfonamide \\(acb\\)$",
  value = "S5 DIURETICS/MASKING AGENTS")
# methylecgonine  ----
# https://www.usada.org/sanction/mike-alexandrov-accepts-doping-sanction/
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "methylecgonine",
  value = "S6 STIMULANTS")
# propylhexadrine  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=propylhexadrine&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "propylhexadrine",
  value = "S6 STIMULANTS")
# androstatrienedione  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=androstatrienedione&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "androstatrienedione",
  value = "S4 HORMONE AND METABOLIC MODULATORS")
# androst-(2,3)-en-17-one  ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=androst&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "androst-\\(2,3\\)-en-17-one",
  value = "S1 ANABOLIC AGENTS")
# methylclostebol ----
# https://www.wada-ama.org/en/prohibited-list?page=0&q=methylclostebol&all=1#search-anchor
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "methylclostebol",
  value = "S1 ANABOLIC AGENTS")
# promagnon  ----
# https://www.usada.org/sanction/u-s-judo-athlete-ohara-accepts-sanction-rule-violation/
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "promagnon",
  value = "S1 ANABOLIC AGENTS")
# 4-hydroxytriamterene  ----
# 4-Hydroxy Triamterene is a diuretic agent and metabolite of Triamterene
# https://www.scbt.com/p/4-hydroxy-triamterene-1226-52-4
tidy_substances <- reclass_substance(
  df = tidy_substances,
  substance = "4-hydroxytriamterene",
  value = "S5 DIURETICS/MASKING AGENTS")
```

After reclassifying the substances above, remaining `UNCLASSIFIED`/`multiple` sanctions are all non-analytic:

```{r}
tidy_substances |>
  dplyr::filter(
      substance_cat == "multiple" & 
      substance_group == "UNCLASSIFIED" & 
      substance_reason != "") |>
  dplyr::distinct(substance_reason) 
```

Changing the `sanction_type` classification to non-analytic requires missing (`NA_character_`) values for `substance_group`:

```{r}
# Change sanction_type to non-analytic
tidy_substances <- tidy_substances |> 
  dplyr::mutate(
    sanction_type = dplyr::case_when(
      substance_group == "UNCLASSIFIED" & substance_reason == "possession" ~ "non-analytic",
      substance_group == "UNCLASSIFIED" & substance_reason == "use/attempted use" ~ "non-analytic",
      substance_group == "UNCLASSIFIED" & substance_reason == "evading sample collection" ~ "non-analytic",
      substance_group == "UNCLASSIFIED" & substance_reason == "non-anatlyical: administration" ~ "non-analytic",
      substance_group == "UNCLASSIFIED" & substance_reason == "trafficking" ~ "non-analytic",
      TRUE ~ sanction_type
    )
  ) 
# Change substance_group to NA_character_
tidy_substances <- tidy_substances |> 
  dplyr::mutate(
    substance_group = dplyr::case_when(
      sanction_type == "non-analytic" ~ NA_character_,
      TRUE ~ substance_group)
    ) 
```

Remove the empty `substance_reason` values: 

```{r}
tidy_substances <- dplyr::filter(tidy_substances, substance_reason != "")
```

## Tidy substances

Now all the substances have been properly classified as [Adverse Analytical Findings](https://www.usada.org/spirit-of-sport/education/alphabet-soup-results-management/#:~:text).

```{r}
tidy_substances |> 
  dplyr::count(sanction_type, substance_group) |> 
  tidyr::pivot_wider(names_from = sanction_type, values_from = n)
```

And the `non-analytic` sanctions are truly [Non-Analytical Anti-doping Rule Violations](https://www.usada.org/spirit-of-sport/education/non-analytical-anti-doping-rule-violations/).

```{r}
dplyr::filter(tidy_substances, sanction_type == "non-analytic") |> 
  dplyr::count(substance_reason, substance_cat) |> 
  tidyr::pivot_wider(names_from = substance_cat, values_from = n)
```


```{r export_data, include=FALSE, echo=FALSE, eval=FALSE}
export_data(x = tidy_substances, path = "../isnt/extdata/demo")
```


