---
title: "Cleaning dates"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Cleaning dates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dopingdata)
```


Below we'll fetch the live USADA sanctions data and process the text:

```{r echo=TRUE, message=FALSE, warning=FALSE}
usada_raw <- get_sanctions_data()
usada <- process_text(raw_data = usada_raw)
```

## Dates

`sanction_announced` contains the date the sanction was announced, and some of these contain two values (`original` and `updated`). Wrangling these values pose some challenges because they aren't *consistently* messy:

```{r}
bad_dates <- subset(usada,
  grepl("^original", usada[['sanction_announced']]),
  c(athlete, sanction_announced))
bad_dates
```

### **clean_dates()**

I've written a `clean_dates()` function that takes `date_col`, `split` and `pattern` arguments:

+ `df` = processed USADA dataset with messy dates

+ `date_col` = sanction date column (usually `sanction_announced`)

+ `split` = regex to pass to split argument of `strsplit()` (defaults to `"updated"`)

+ `pattern` = regex for other non-date pattern (defaults to `"original"`)

Below, `clean_dates()` is demonstrated on a couple of the messy rows found above:

```{r}
clean_dates(
  df = head(bad_dates, 2),
  date_col = "sanction_announced",
  split = "updated",
  pattern = "original")
```

For `usada`, split the data into three `data.frame`s (`bad_dates`, `good_dates`, and `no_dates`).

```{r}
bad_dates <- subset(usada, 
  grepl("^original", usada[['sanction_announced']]))
good_dates <- subset(usada, 
  !grepl("^original", usada[['sanction_announced']]) & sanction_announced != "")
no_dates <- subset(usada,
  athlete == "*name removed" & sanction_announced == "")
```

Clean dates in `bad_dates` by splitting the bad dates on `"updated"` and provided `"original"` as the pattern (the opposite will also work). The `sanction_date` column will contain the correctly formatted updated `sanction_date`.

After formatting `good_dates` and removing `original_date` column we can combine the two with `rbind()`.

```{r}
cleaned_dates <- clean_dates(
  df = bad_dates, 
  date_col = "sanction_announced", 
  split = "updated", 
  pattern = "original")
# address names 
names(cleaned_dates)[names(cleaned_dates) == 'split_date'] <- 'sanction_date'
names(cleaned_dates)[names(cleaned_dates) == 'pattern_date'] <- 'original_date'
# format good_dates
good_dates$sanction_date <- as.Date(x = good_dates[['sanction_announced']], 
                                    format = "%m/%d/%Y")
# get intersecting names 
nms <- intersect(names(cleaned_dates), names(good_dates))
# bind the two datasets 
usada_dates <- rbind(good_dates, cleaned_dates[nms])
str(usada_dates)
```


