---
title: "clean-sports"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{clean-sports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r echo=TRUE, message=FALSE, warning=FALSE}
usada_dates <- read.csv("../data-raw/csv/2023-12-21/2023-12-21-usada_dates.csv")
```

```{r setup}
library(dopingdata)
library(dplyr)
library(stringr)
library(tidyr)
library(forcats)
```


## Sports 

To wrangle the sports, I'll use packages and functions from the `tidyverse` (`dplyr`, `stringr`, `tidyr`, etc.), but I also provide the base R alternatives (wherever possible).

```{r}
library(tidyverse)
```

`tidyverse` functions will return a `tibble` (not a `data.frame`), which prints fewer rows to the console. 

```{r}
usada_sports <- tibble::as_tibble(usada_dates)
str(usada_sports)
```

We can start by counting the `sport` column:

```{r count-sports-raw}
usada_sports |> 
  dplyr::count(sport, sort = TRUE)
```

### Support personnel

Some of the sports aren't sports--they're `athlete support personnel`. These need a  `support_personnel` identifier.

```{r support_personnel}
usada_sports <- dplyr::mutate(usada_sports, 
  # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(
      sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA)) 

usada_sports |> 
  dplyr::filter(stringr::str_detect(sport, "support personnel")) |> 
  dplyr::count(sport, support_personnel) |> 
  tidyr::pivot_wider(names_from = support_personnel, values_from = n)
```

### 'track and field' or 'track & field'

Convert sports like `track & field` to `track and field` to help determine which athletes/support personnel are involved in multiple sports.

```{r track-field}
usada_sports <- dplyr::mutate(usada_sports,
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field'))

usada_sports |> 
  dplyr::filter(stringr::str_detect(sport, "track")) |> 
  dplyr::count(sport, support_personnel) |> 
  tidyr::pivot_wider(names_from = support_personnel, values_from = n)
```


### Spelling 

The incorrect spelling for `brazilian jiu-jitsu` (`brazillian jiu-jitsu`) is corrected below.

```{r bjj}
usada_sports <- dplyr::mutate(usada_sports, 
  # brazilian jiu-jitsu
  sport = dplyr::case_when(
    sport == 'brazillian jiu-jitsu' ~ 'brazilian jiu-jitsu',
    TRUE ~ sport)) 
usada_sports |> 
  dplyr::filter(stringr::str_detect(sport, "jitsu")) |> 
  dplyr::count(sport, sort = TRUE)
```


### 'paralympic'

An identifier for paralympic sports: `paralympic`. 

```{r paralympic}
usada_sports <- dplyr::mutate(usada_sports, 
  # paralympic
  paralympic = 
    dplyr::if_else(condition = stringr::str_detect(sport, "paralympic|para"), 
      true = TRUE, false = FALSE, missing = NA)) 

usada_sports |> 
  dplyr::filter(stringr::str_detect(sport, "paralympic|para")) |> 
  dplyr::count(paralympic, sport) |> 
  tidyr::pivot_wider(names_from = paralympic, values_from = n)
```

### Multiple sports

Identify the multiple sports using `and` and `, ` in a regular expression. 

```{r multiple_sports}
usada_sports <- dplyr::mutate(usada_sports, 
  # multiple_sports
  multiple_sports = 
    if_else(condition = stringr::str_detect(sport, "and |, "), 
      true = TRUE, false = FALSE, missing = NA))

usada_sports |> 
  dplyr::filter(stringr::str_detect(sport, "and |, ")) |> 
  dplyr::count(multiple_sports, sport) |> 
  tidyr::pivot_wider(names_from = multiple_sports, values_from = n)
```

### Tidy

Separate the multi-sport athletes in `usada_sports` as `multp_sport_athletes` and single-sport athletes in `single_sport_athletes`.

```{r create-multp_sport_athletes-single_sport_athletes, collapse=TRUE}
multp_sport_athletes <- usada_sports |> 
  dplyr::filter(multiple_sports == TRUE)
str(multp_sport_athletes)
single_sport_athletes <- usada_sports |> 
  dplyr::filter(multiple_sports == FALSE)
str(single_sport_athletes)
```


The athletes listed with multiple sports will occupy multiple rows in a 'tidy' version of `usada_sports`. 

  -   Passing the sport column to `tidyr::separate_rows()` and `stringr::str_trim()` in  `multp_sport_athletes` will create a `tidy_multp_sport_athletes` dataset:

```{r tidy_multp_sport_athletes}
tidy_multp_sport_athletes <- multp_sport_athletes |> 
  tidyr::separate_rows(sport, sep = "and|, ") |> 
  dplyr::mutate(sport = stringr::str_trim(sport, side = "both"))
str(tidy_multp_sport_athletes)
```

Finally, combine the two datasets.

```{r tidy_sports}
tidy_sports <- dplyr::bind_rows(single_sport_athletes, tidy_multp_sport_athletes)
str(tidy_sports)
```

### **clean_sports()**

These steps are combined in the `clean_sports()` function: 

```{r clean_sports-function, eval=TRUE}
str(
  clean_sports(
    df = usada_dates, 
    sport_col = "sport", 
    tidy = TRUE)
)
```

Verify there aren't any duplicates (again). 

```{r}
tidy_sports |> 
  dplyr::count(athlete, sanction_date, sport) |> 
  dplyr::filter(n > 1)
```

We can see the multi-sport athletes are listed in `tidy_sports` (but with one sport per row):

```{r check-tidy_sports}
tidy_sports |> 
    dplyr::filter(multiple_sports == TRUE) |> 
    dplyr::select(athlete, sport)
```



