---
title: "Multiple AAF substances"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Multiple AAF substances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(
  pillar.print_max = 25, 
  dplyr.print_max = 25,
  knitr.kable.NA = '',
  pillar.print_min = 25)
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

```{r pkg}
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

## Motivation 

This vignette covers the classification of adverse analytic findings (AAF) for multiple banned substances in the [USADA sanction table](https://www.usada.org/testing/results/sanctions/). 

## Data versioning 

`dopingdata` uses data from tables that get updated regularly. To keep track of the versions, I've separated the raw (`_raw`) and processed (`_pro`) data in the `inst/` folder:

  - Raw data:
    ```{r raw-tree, echo=FALSE}
    fs::dir_tree("../inst/extdata/raw/", recurse = FALSE)
    ```

  - Processed data
    ```{r extdata-tree, echo=FALSE}
    fs::dir_tree("../inst/extdata", recurse = FALSE)
    ```

## Outline 

This vignette assumes the following: 

1. **Harvest the data:** A copy of the [USADA sanctions table](https://www.usada.org/testing/results/sanctions/) has been scraped and lives in the `inst/extdata/raw/` folder 

  - See [`vignettes/scraping-usada.Rmd`](https://mjfrigaard.github.io/dopingdata/articles/scraping-usada-sanctions.html) for more information
  
    ```{r raw-usada_sanctions, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/raw/", regexp = "sanctions"
    )
    ```

  - These data have had all the column names formatted with `janitor::clean_names()`, and all the text has been converted to lowercase 
    ```{r pro-usada_sanctions, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/", regexp = "sanctions"
    )
    ```

2. **Fix the dates:** The dates in this table have been formatted and wrangled 

  - See [`vignettes/sanction-dates.Rmd`](https://mjfrigaard.github.io/dopingdata/articles/sanction-dates.html) more information 
  
    ```{r usada_dates, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/",
      regexp = "dates")
    ```

3. **Tidy the sports:** The `sport` column has been 'tidied' and athletes/support personnel listed with more than one sport have been repeated in the data  

  - See [`vignettes/sanction-sports.Rmd`](https://mjfrigaard.github.io/dopingdata/articles/sanction-sports.html) for more information  
  
    ```{r tidy_usada_sports, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/", 
      regexp = "sports")
    ```
    
4. **Tidy the AAF (single substances):** The `sanction_reason` column in `tidy_sports.csv` was used to classify the analytic (i.e, adverse analytic finding) sanctions into `sanction_groups` based on the WADA banned list of substances. 

  - See [`vignettes/single-aaf-substances.Rmd`](https://mjfrigaard.github.io/dopingdata/articles/single-aaf-substances.html) for more information  
  
    ```{r tidy_single_aafs, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/", 
      regexp = "single_aafs")
    ```

### Data 

This vignette picks up where [`vignettes/single-aaf-substances.Rmd`](https://mjfrigaard.github.io/dopingdata/articles/single-aaf-substances.html) left off (and imports the previous `tidy_single_aafs.csv` dataset)

```{r get_recent, eval=FALSE}
get_recent("../inst/extdata", "csv")
✔ The last modified date in '../inst/extdata': 2023-01-23
       dstmp 
"2023-01-23" 
```

```{r import-substance_types}
substance_types <- readr::read_csv(
  file = '../inst/extdata/usada_substance_types.csv')
dplyr::glimpse(substance_types)
```

### Multiple substances (`multiple_sr`)

The substance_types data has the `multiple_sr` column for identifying if an sanction included more than one substance:

```{r multiple_sr-tabyl}
substance_types |> 
  janitor::tabyl(multiple_sr) |> 
  janitor::adorn_pct_formatting(digits = 0)
```

## AAF vs. ADRV Sanctions

The sanctions are divided into two categories: 

- `analytic`: [Adverse Analytical Finding](https://www.usada.org/spirit-of-sport/education/alphabet-soup-results-management/#:~:text), AAF; *An AAF is a report from a WADA-accredited laboratory that identifies the presence of a prohibited substance and/or its metabolites or markers in a sample.* 

- `non-analytic`: [Non-Analytical Anti-doping Rule Violation](https://www.usada.org/spirit-of-sport/education/non-analytical-anti-doping-rule-violations/) ADRV; *a non-analytical anti-doping rule violation does not stem from a positive urine or blood sample, but instead originates from, and is substantiated by, other evidence of doping or violations by an athlete or athlete support personnel.*. 


The variable identifying the AAF vs. ADRV sanctions is `sanction_type`, and I breakdown  `multiple_sr` by `sanction_type` below: 

```{r multiple_sr-sanction_type-tabyl}
substance_types |> 
  janitor::tabyl(multiple_sr, sanction_type) |> 
  janitor::adorn_percentages(denominator = "col") |> 
  adorn_pct_formatting(digits = 0) 
```

### Multiple analytic substances

Filter `substance_types` into multiple `analytic` substances (`mult_aaf_subs`)

```{r create-mult_aaf_subs}
mult_aaf_subs <- substance_types |>
  dplyr::filter(sanction_type == "analytical" & multiple_sr == TRUE)
mult_aaf_subs |> 
  dplyr::count(multiple_sr)
```

Below is a 'count of counts', which tells me most of the multiple substance combinations occur only once. 

```{r count-substance_reason}
mult_aaf_subs |>
  dplyr::count(substance_reason, sort = TRUE, name = "Count") |>
  dplyr::count(Count, sort = TRUE)
```

## Tidying multiple substances

We're going to separate the multiple [WADA banned substances](https://www.wada-ama.org/en/prohibited-list?q) into a 'tidy' form (one substance per athlete per row), which means the athletes with multiple substances will appear in this dataset more than once.

To tidy the substances with multiple steps, I'll define a regular expression that identifies the separators between each substance (and make sure we're not splitting up the individual substances). The regular expression should (ideally) cover a range of word boundaries, tabs, semicolons, and spaces. 

### Step 1: white space 

I'll start by trimming the white space from `substance_reason`

```{r trim-white-space}
dplyr::mutate(mult_aaf_subs,
  substance_reason = stringr::str_trim(substance_reason, "both")) |> 
  dplyr::select(athlete, substance_reason)
```

### Step 2: commas in chemical compounds 

I need to be careful using commas, because some of the chemical compounds use commas in their names:

```{r comma-chemical-substances}
stringr::str_view_all(
  mult_aaf_subs$substance_reason,
  pattern = "\\d+,\\d+",
  match = TRUE
)
```

### Step 3: plurals

The pattern above seems to identify most of the separators, but it's separating on `<substance>` **`and`** `its metabolite`:

```{r view-metabolite-metabolites}
# metabolite/metabolites
stringr::str_view_all(
  mult_aaf_subs$substance_reason,
  "and its metabolites|and its metabolite|its metabolites|its metabolite",
  match = TRUE
)
```


I want to include the information on `metabolite`/`metabolites`, so I'll replace this with, `(metabolite)`:

```{r metabolite-metabolites}
dplyr::mutate(mult_aaf_subs,
  substance_reason = stringr::str_replace_all(
    substance_reason,
    "and its metabolite|and its metabolites",
    "(metabolite)"),
  substance_reason = stringr::str_replace_all(
    substance_reason,
    "its metabolite",
    "(metabolite)")) |>
  dplyr::select(substance_reason) |> 
  purrr::as_vector() |> 
  stringr::str_view_all(pattern = "metabolite", match = TRUE)
```

### Step 4: conjunctions

I want to create a regular expression that matches on the conjunctions (`and` and `,`) connecting the multiple substances. I'll store the pattern in `multp_subs_regex`.

```{r view-and-comma-boundaries}
# I'm showing only a sample of 20 to reduce printing
multp_subs_regex <- "\\b and \\b|\\b, and \\b|, and |\\band\\b|\\b, \\b|, | & |\\b&\\b"
stringr::str_view_all(
  sample(mult_aaf_subs$substance_reason, 
    size = 20, replace = FALSE),
  pattern = multp_subs_regex,
  match = TRUE
)
```

### Step 5: semicolons

In addition to conjunctions, I also want to match on the semicolons, so I will add this to `multp_subs_regex`. 

```{r view-semicolon-boundaries}
# I'm showing only a sample of 20 to reduce printing
stringr::str_view_all(
  sample(mult_aaf_subs$substance_reason,
    size = 20,
    replace = FALSE
  ),
  pattern = ";\\b|; \\b|; \\b |; ",
  match = TRUE
)
# add to multp_subs_regex
multp_subs_regex <-
  paste0(multp_subs_regex, ";\\b|; \\b|; \\b |; ", collapse = "|")
```

### Step 6: tab boundaries

The final pattern to add to `multp_subs_regex` is for tab boundaries (`\\t`). 

```{r view-tab-boundaries}
stringr::str_view_all(
    mult_aaf_subs$substance_reason,
  pattern = "\\t| \\t| \\t |\\t ",
  match = TRUE)
# add to multp_subs_regex
multp_subs_regex <- paste0(multp_subs_regex, "\\t| \\t| \\t |\\t ", collapse = "|")
```

I combine all the steps above into the code chunk below: 

```{r clean-mult_aaf_subs-substance_reason}
# white space
mult_aaf_subs <- dplyr::mutate(mult_aaf_subs,
  substance_reason = stringr::str_trim(
    substance_reason, 
    "both"),
  # plurals
    substance_reason = stringr::str_replace_all(
    substance_reason,
    "and its metabolite|and its metabolites",
    "(metabolite)"),
  # plurals
  substance_reason = stringr::str_replace_all(
    substance_reason,
    "its metabolite",
    "(metabolite)"),
  # regex
  substance_reason = stringr::str_remove_all(
    substance_reason,
    multp_subs_regex
  ))
```

## Classifying multiple substances 

I've written `classify_wada_substances()` to identify sanctions with multiple [WADA banned substances](https://www.wada-ama.org/en/prohibited-list#search-anchor).

The `classify_wada_substances()` function scans the `substance_reason` column and identifies any substances found on the [WADA list](https://www.wada-ama.org/sites/default/files/2022-09/2023list_en_final_9_september_2022.pdf). 

See [`vignettes/single-aaf-substances.Rmd`](https://mjfrigaard.github.io/dopingdata/articles/single-aaf-substances.html) for more information.

### `tidy_multp_aaf_subs`

The output from is stored in `tidy_multp_aaf_subs`:

```{r tidy_multp_aaf_subs}
tidy_multp_aaf_subs <- classify_wada_substances(
  usada_data = mult_aaf_subs,
  subs_column = substance_reason
)
tidy_multp_aaf_subs |>
  janitor::tabyl(substance_group) |> 
  janitor::adorn_pct_formatting()
```

### Unclassified substances 

I want to identify any unclassified substances in `tidy_multp_aaf_subs`:

```{r unclassified_multp_analytic_substance}
tidy_multp_aaf_subs |>
  dplyr::filter(is.na(substance_group)) |>
  dplyr::count(substance_reason, sort = TRUE)
```

I can see this particular `substance_reason` is non-analytical, so I need to reclassify this in the `usada_substance_types.csv` data (which we imported above as `substance_types`). 

We can identify this sanction and athlete below:

```{r identify-reclass}
dplyr::filter(tidy_multp_aaf_subs, 
  stringr::str_detect(substance_reason, "^non-anatlyical")) |> 
  dplyr::select(athlete, substance_reason)
```

I can recode these in `substance_types` and `tidy_multp_aaf_subs`:

```{r recode-non-analytical}
reclass_substance_types <- dplyr::mutate(substance_types,
  sanction_type = dplyr::case_when(
    stringr::str_detect(
      substance_reason, "non-anatlyical") ~ "non-analytical",
    TRUE ~ sanction_type
  )) 

tidy_multp_aaf_subs <- dplyr::mutate(tidy_multp_aaf_subs,
  sanction_type = dplyr::case_when(
    stringr::str_detect(
      substance_reason, "non-anatlyical") ~ "non-analytical",
    TRUE ~ sanction_type
  ))
```

Check the `sanction_type` in `reclass_substance_types`

```{r reclass_substance_types-sanction_type}
reclass_substance_types |> 
  dplyr::filter(stringr::str_detect(athlete, "block, mark")) |> 
  dplyr::select(athlete, substance_reason, sanction_type)
```

Check the `sanction_type` in `tidy_multp_aaf_subs`

```{r check-reclass-sanction_type-tidy_multp_aaf_subs}
tidy_multp_aaf_subs |> 
  dplyr::filter(stringr::str_detect(athlete, "block, mark")) |> 
  dplyr::select(athlete, substance_reason, sanction_type)
```

## Export substance types

Export `substance_types`

```{r export_extdata-substance_types}
export_extdata(
  x = substance_types, 
  inst_path = "../inst/", 
  raw = FALSE
)
```


## Export tidy classified (multiple) substances 

Export classified multiple substances (`tidy_multp_aaf_subs`):

```{r export_extdata-tidy_multp_aaf_subs}
export_extdata(
  x = tidy_multp_aaf_subs, 
  inst_path = "../inst/", 
  raw = FALSE
)
```

