---
title: "Sanction sports"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Sanction sports}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(
  pillar.print_max = 25, 
  dplyr.print_max = 25,
  knitr.kable.NA = '',
  pillar.print_min = 25)
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

Packages:

```{r pkg, message=FALSE, warning=FALSE}
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

## Motivation 

This vignette covers the `sports` listed in the [USADA sanctions table](https://www.usada.org/testing/results/sanctions/). 

## Data versioning 

This package uses data from a table that gets updated regularly. To keep track of the versions, I've organized the `inst/extdata` folder by date (`YYYY-MM-DD`), but I have a helper function for getting the recent version:

```{r tdy_stmp}
tdy_stmp <- get_recent_version("../inst/extdata/raw")
tdy_stmp
```

Now I can use `tdy_stmp` whenever I need to use find the data version! See below:


## Outline 

This vignette assumes the following: 

1. **Scrape the website:** A copy of the [USADA sanctions table](https://www.usada.org/testing/results/sanctions/) has been scraped and lives in the `inst/extdata/raw/` and `inst/extdata/pro/` folders 

  - See the `scrape_usada_sanctions()` function and `scraping-usada-sanctions.Rmd` vignette for more information
  
    ```{r raw-usada-sanctions, echo=FALSE}
    fs::dir_tree(paste0("../inst/extdata/raw/", tdy_stmp), regexp = "sanctions")
    ```

   - These data have had all the column names formatted with `janitor::clean_names()`, and all the text has been converted to lowercase 
   
    ```{r pro-usada-sanctions, echo=FALSE}
    fs::dir_tree(paste0("../inst/extdata/pro/", tdy_stmp), regexp = "sanctions")
    ```
    
2. **Fix the dates:** The dates in this table have been formatted and wrangled 

  - See the  `vignettes/sanction-dates.Rmd` vignette for more information 
  
    ```{r usada_dates, echo=FALSE}
    fs::dir_tree(
      path = "../inst/extdata/der/", 
      regexp = "dates")
    ```

### Data 

I'm going to locate the most recently wrangled data file from the `inst/extdata/pro/` folder:

``` r
get_recent_data_file(
  folder = paste0("../inst/extdata/der/", tdy_stmp), 
  type = "csv")
✔ import code pasted to clipboard!
✔ use: readr::read_csv('../inst/extdata/der/sanction_dates.csv')
```

The path on the clipboard is the exported data from `vignettes/sanction-dates.Rmd` (if not, please head over to that vignette and fix the dates before moving forward).


```{r sanctions}
sanctions <- readr::read_csv(
  file = '../inst/extdata/der/sanction_dates.csv')
dplyr::glimpse(sanctions)
```


## `sport` counts (raw)

```{r count-sports-raw}
sanctions |> 
  dplyr::count(sport, sort = TRUE)
```

### `athlete support personnel`

I'll create an identifier for athlete support personnel, `support_personnel`. 

```{r support_personnel}
dplyr::mutate(sanctions, 
  # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA)) |> 
  dplyr::count(sport, support_personnel) |> 
  tidyr::pivot_wider(names_from = support_personnel, values_from = n)
```

### `track and field` 

I'm going to convert `track and field` as `track & field` (this will help me  determine which athletes/support personnel are involved in multiple sports). 

```{r track-field}
dplyr::mutate(sanctions,
  # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA),
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field')) |> 
  dplyr::count(sport, sort = TRUE)
```


### `brazilian jiu-jitsu`

There is an incorrect spelling for `brazilian jiu-jitsu` (`brazillian jiu-jitsu`), which I will correct here: 

```{r bjj}
dplyr::mutate(sanctions, 
  # # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA),
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field'),
  # brazilian jiu-jitsu
  sport = dplyr::case_when(
    sport == 'brazillian jiu-jitsu' ~ 'brazilian jiu-jitsu',
    TRUE ~ sport)) |> 
  dplyr::count(sport, sort = TRUE)
```


### `paralympic`

I'll also create an identifier for paralympic sports, `paralympic`. 

```{r paralympic}
dplyr::mutate(sanctions, 
  # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA),
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field'),
  # brazilian jiu-jitsu
  sport = dplyr::case_when(
    sport == 'brazillian jiu-jitsu' ~ 'brazilian jiu-jitsu',
    TRUE ~ sport),
  # paralympic
  paralympic = 
    dplyr::if_else(condition = stringr::str_detect(sport, "paralympic|para"), 
      true = TRUE, false = FALSE, missing = NA)) |> 
  dplyr::count(paralympic, sport) |> 
  tidyr::pivot_wider(names_from = paralympic, values_from = n)
```

### `multiple_sports` 

Now I can identify the multiple sports using `and` and `, ` in the regular expression. 

```{r multiple_sports}
dplyr::mutate(sanctions, 
  # support_personnel
  support_personnel = 
    if_else(condition = stringr::str_detect(sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA),
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field'),
  # brazilian jiu-jitsu
  sport = case_when(
    sport == 'brazillian jiu-jitsu' ~ 'brazilian jiu-jitsu',
    TRUE ~ sport),
  # multiple_sports
  multiple_sports = 
    if_else(condition = stringr::str_detect(sport, "and |, "), 
      true = TRUE, false = FALSE, missing = NA),
  # paralympic
  paralympic = 
    if_else(condition = stringr::str_detect(sport, "paralympic|para"), 
      true = TRUE, false = FALSE, missing = NA)) |> 
  dplyr::count(multiple_sports, sport) |> 
  tidyr::pivot_wider(names_from = multiple_sports, values_from = n)
```

## `sanctions_sports`

I'll store these new variables in `sanctions_sports`

```{r sanctions_sports}
sanctions_sports <- dplyr::mutate(sanctions, 
  # support_personnel
  support_personnel = 
    dplyr::if_else(condition = stringr::str_detect(sport, "support personnel"), 
      true = TRUE, false = FALSE, missing = NA),
  # track & field
  sport = stringr::str_replace_all(sport, 'track and field', 'track & field'),
  # brazilian jiu-jitsu
  sport = case_when(
    sport == 'brazillian jiu-jitsu' ~ 'brazilian jiu-jitsu',
    TRUE ~ sport),
  # multiple_sports
  multiple_sports = 
    dplyr::if_else(condition = stringr::str_detect(sport, "and |, "), 
      true = TRUE, false = FALSE, missing = NA),
  # paralympic
  paralympic = 
    dplyr::if_else(condition = stringr::str_detect(sport, "paralympic|para"), 
      true = TRUE, false = FALSE, missing = NA)) 
dplyr::glimpse(sanctions_sports)
```

### Multiple sports 

Some athletes were listed with multiple sports, which means they'll occupy multiple rows in a 'tidy' version of this dataset. 

We can check this using `count()` below to verify we have one athlete for each sport:

```{r check-tidy-sports}
sanctions_sports |> 
  dplyr::filter(multiple_sports == TRUE) |> 
  tidyr::separate_rows(sport, sep = "and|, ") |> 
  dplyr::count(athlete, sport)
```

### `tidy_multp_sports`

I'll remove any white-space left over from `tidyr::separate_rows()` with `stringr::str_trim()` and create `tidy_multp_sports`. 

```{r tidy_multp_sports}
tidy_multp_sports <- sanctions_sports |> 
  dplyr::filter(multiple_sports == TRUE) |> 
  tidyr::separate_rows(sport, sep = "and|, ") |> 
  dplyr::mutate(sport = stringr::str_trim(sport, side = "both"))
glimpse(tidy_multp_sports)
```

### `single_usada_sports`

I need to remove the single occurrences of these athletes in `sanctions_sports` and replace them with the tidy athlete names. 

```{r single_usada_sports}
multp_sports_athletes <- unique(tidy_multp_sports['athlete'])
single_usada_sports <- dplyr::filter(sanctions_sports, 
                                        athlete %nin% multp_sports_athletes)
```

Verify there aren't any duplicates in `single_usada_sports` (there should be zero rows here)

```{r check-single_usada_sports}
single_usada_sports |> 
  dplyr::count(athlete, sanction_date, sport) |> 
  dplyr::filter(n > 1)
```

## `tidy_sports`

Now combine these into `tidy_sports` and verify there aren't any duplicates (again). 

```{r tidy_sports}
tidy_sports <- dplyr::bind_rows(single_usada_sports, tidy_multp_sports)
tidy_sports |> 
  dplyr::count(athlete, sanction_date, sport) |> 
  dplyr::filter(n > 1)
```

```{r}
dplyr::glimpse(tidy_sports)
```


Ready to export!

```{r vroom_write-tidy_sports}
new_pth <- "../inst/extdata/der/"
fs::dir_create(new_pth)
csv_pth <- paste0(new_pth, "tidy_sports.csv")
csv_pth
```

Export `tidy_sports`

```{r export-sports, eval=FALSE}
vroom::vroom_write(x = tidy_sports, file = csv_pth, delim = ",")
```

Verify 

```{r}
fs::dir_tree(new_pth, regexp = "tidy_sports|dates")
```

***CHECK:***  Can I can see the `dates` and `sports` data in the data folder?
