---
title: "Sanction dates"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Sanction dates}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(
  pillar.print_max = 25, 
  dplyr.print_max = 25,
  knitr.kable.NA = '',
  pillar.print_min = 25)
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

Packages:

```{r pkg, message=FALSE, warning=FALSE}
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

# Motivation 

This vignette covers how format and wrangle the sanction dates in the [USADA sanctions table](https://www.usada.org/testing/results/sanctions/). 

## Data versioning 

This package uses data from a table that gets updated regularly. To keep track of the versions, I've organized the `inst/extdata` folder by date (`YYYY-MM-DD`), but I have a helper function for getting the recent version:

```{r tdy_stmp}
tdy_stmp <- get_recent_version("../inst/extdata/raw")
tdy_stmp
```

Now I can use `tdy_stmp` whenever I need to use find the data version! See below:

## Outline 

This vignette assumes the following: 

1. **Scrape the website:** A copy of the [USADA sanctions table](https://www.usada.org/testing/results/sanctions/) has been scraped and lives in the `inst/extdata/raw/` folder 

  - See the `scrape_usada_sanctions()` function and `scraping-usada.Rmd` vignette for more information
  
    ```{r raw-usada_data, echo=FALSE}
    fs::dir_tree(paste0("../inst/extdata/raw/", tdy_stmp))
    ```

   - These data have had all the column names formatted with `janitor::clean_names()`, and all the text has been converted to lowercase 


## Import recent data file 

I've written a custom import function, `get_recent_data_file()`, which helps with locating/importing the most recent USADA data:

``` r
get_recent_data_file(
  folder = paste0("../inst/extdata/pro/", tdy_stmp), type = "csv")
✔ import code pasted to clipboard!
✔ use: readr::read_csv('../inst/extdata/pro/2023-01-20/2023-01-20-usada_sanctions_pro.csv')
```

```{r usada_raw}
usada_raw <- readr::read_csv(
  '../inst/extdata/pro/2023-01-19/2023-01-19-usada_sanctions_pro.csv')
dplyr::glimpse(usada_raw)
```


### Sanction dates

I need to format the dates in `sanction_announced` as an actual date, but should check to see if they're all in the same format. 

```{r check-dates}
dplyr::mutate(usada_raw, 
  sanction_announced = lubridate::mdy(sanction_announced)) |> 
  utils::head(25)
```

Which ones failed?  Store these in `bad_dates`.

```{r bad_dates}
bad_dates <- mutate(usada_raw, 
  sanction_date = lubridate::mdy(sanction_announced)) |> 
  dplyr::filter(is.na(sanction_date) & !is.na(sanction_announced)) 
bad_dates |> 
  dplyr::select(sanction_announced) |> 
  head(10)
```

I can see they've stuffed two dates in the `sanction_announced` column for 29 athletes: 

1. I will use `tidyr::separate_rows()` to split the dates into `original` and `updated`  
2. Remove the excess colons and white-space with `stringr::str_remove_all()`  
3. `filter` out the original dates (I only want the updated date)  
4. convert the dates to date with `anytime::anydate()`  
5. take care of the single outlier (formatted with a two-year date)  
6. remove the `sanction_announced` column (use `sanction_date` instead)  


```{r good_dates}
good_dates <- tidyr::separate_rows(data = bad_dates,
  sanction_announced, sep = "updated") |>
  dplyr::mutate(sanction_date = stringr::str_remove_all(sanction_announced, ": ")) |>
  dplyr::filter(!stringr::str_detect(sanction_date, "^original")) |>
  dplyr::mutate(sanction_date = anytime::anydate(sanction_date)) |>
  dplyr::mutate(
    sanction_date = dplyr::case_when(
      athlete == "ngetich, eliud" ~ lubridate::as_date("2022-01-25"),
      TRUE ~ sanction_date
    )
  ) |>
  dplyr::select(-sanction_announced)
good_dates |> glimpse()
```

Now I want to format the dates in `usada_raw` using `anytime::anydate()`, then filter out the missing `sanction_date`s that didn't parse, remove the `sanction_announced` column, and bind this back to the `good_dates` table with fixed sanction dates.  

```{r usada_dates}
usada_dates <- dplyr::mutate(usada_raw, 
    sanction_date = anytime::anydate(sanction_announced)) |> 
  dplyr::filter(!is.na(sanction_date)) |> 
  dplyr::select(-sanction_announced) |> 
  dplyr::bind_rows(good_dates) 
dplyr::glimpse(usada_dates)
```


## Export 

Create paths for today's data:

```{r paths-dates}
new_pth <- paste0("../inst/extdata/der/")
fs::dir_create(new_pth)
csv_pth <- paste0(new_pth, "sanction_dates.csv")
csv_pth
```

Export the `usada_dates` as .csv and .rds

```{r export-dates, eval=FALSE}
vroom::vroom_write(x = usada_dates, file = csv_pth, delim = ",")
```

Verify

```{r verify-export}
fs::dir_tree(new_pth, regexp = "dates")
```


