---
title: "Scraping USADA sanctions"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 4
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Scraping USADA sanctions}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  eval = FALSE,
  collapse = TRUE,
  comment = "#>"
)
options(
  pillar.print_max = 25,
  dplyr.print_max = 25,
  knitr.kable.NA = '',
  pillar.print_min = 25,
  repos = "https://cloud.r-project.org")
library(dopingdata)
library(robotstxt)
options(robotstxt_warn = FALSE)
```

Packages:

```{r pkg, eval=TRUE, message=FALSE, warning=FALSE}
library(dopingdata)
library(robotstxt)
options(robotstxt_warn = FALSE)
```

## USADA sanction data

The data comes from the [USADA sanctions page](https://www.usada.org/results/sanctions/), which
hosts two tables:

- the **sanctions table** (Athlete, Sport, Substance/Reason, Sanction Terms, Sanction Announced)
- the **prohibited association table**, listing athlete support personnel (coaches, trainers,
  etc.) currently ineligible under the Prohibited Association rule (Name, Suspension Ends)

## Use your manners

Because this package is built on top of the efforts of the fine people who collected, organized,
and shared their data, `dopingdata` uses the [`polite` package](https://dmi3kno.github.io/polite/)
to fetch this page, which respects `robots.txt` and its crawl-delay rules.

### Check `robots.txt`

```{r rt}
rtxt <- robotstxt::robotstxt(domain = "https://www.usada.org/")

rtxt$check(
  paths = c("results/", "results/sanctions/"),
  bot   = "*"
)
```

## Getting the data

Rather than scraping the page yourself, `dopingdata` provides a small cached API modeled on the
[`fightr`](https://benyamindsmith.github.io/fightr/index.html) package: `get_*()` functions
return data from a local cache and refresh it automatically once it's more than `max_age_days`
old (7 days by default), and `update_*()` functions force a refresh. Fetching only ever happens
when you explicitly call one of these functions — never on package load.

```{r get_sanctions_data, eval=FALSE}
sanctions <- get_sanctions_data()
str(sanctions)
```

```{r get_prohibited_association, eval=FALSE}
prohibited_association <- get_prohibited_association()
str(prohibited_association)
```

To force a refresh regardless of cache age:

```{r update_sanctions_data, eval=FALSE}
update_sanctions_data(verbose = TRUE)
```

`update_sanctions_data()` makes a single request for the page and refreshes both cached tables,
so calling `get_sanctions_data()` and `get_prohibited_association()` back to back doesn't trigger
two separate scrapes.

## Exporting the data

Some common tasks (like exporting a data frame as a `.csv` file into a date-stamped folder) have
been wrapped in functions:

```{r usada_sanctions_raw, echo=TRUE, message=FALSE, warning=FALSE, eval=TRUE}
usada_sanctions_raw <- get_sanctions_data()
str(usada_sanctions_raw)
```

```{r eval=FALSE}
export_data(
  x = usada_sanctions_raw,
  path = "../dev")
```

There's also an `export_extdata()` function if you're storing the data in a package:

```{r eval=FALSE}
export_extdata(
  x = usada_sanctions_raw,
  path = "dev")
```
