---
title: "Single AAF substances"
output: 
  rmarkdown::html_vignette:
    df_print: kable
vignette: >
  %\VignetteIndexEntry{Single AAF substances}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
options(
  pillar.print_max = 25, 
  dplyr.print_max = 25,
  knitr.kable.NA = '',
  pillar.print_min = 25)
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

Packages:

```{r pkg, message=FALSE, warning=FALSE}
library(dopingdata)
library(dplyr)
library(tidyr)
library(stringr)
library(forcats)
library(lubridate)
library(readr)
library(readxl)
library(janitor)
library(anytime)
```

## Motivation 

This vignette covers the classification of adverse analytic findings for a single banned substance in the [USADA sanction table](https://www.usada.org/testing/results/sanctions/). 

## Data versioning 

`dopingdata` uses data from tables that get updated regularly. To keep track of the versions, I've separated the raw (`_raw`) and processed (`_pro`) data in the `inst/` folder: 

```{r}
fs::dir_tree("../inst/extdata", recurse = FALSE)
```




by date (`YYYY-MM-DD`), but I also have a helper function for getting the recent version:

```{r tdy_stmp}
get_recent("../inst/extdata/", full = TRUE)
get_recent("../inst/extdata/raw", full = TRUE)
```


## Outline 

This vignette assumes the following: 

1. **Harvest the data:** A copy of the [USADA sanctions table](https://www.usada.org/testing/results/sanctions/) has been scraped and lives in the `inst/extdata/raw/` folder 

  - See the `scrape_usada_sanctions()` function and `scraping-usada.Rmd` vignette for more information
  
    ```{r raw-usada_sanctions, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/raw/", regexp = "sanctions"
    )
    ```

  - These data have had all the column names formatted with `janitor::clean_names()`, and all the text has been converted to lowercase 
    ```{r pro-usada_sanctions, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/", regexp = "sanctions"
    )
    ```

2. **Fix the dates:** The dates in this table have been formatted and wrangled 

  - See the  `vignettes/sanction-dates.Rmd` vignette for more information 
  
    ```{r usada_dates, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/",
      regexp = "dates")
    ```

3. **Tidy the sports:** The `sport` column has been 'tidied' and athletes/support personnel listed with more than one sport have been repeated in the data  

  - See the `vignettes/sports.Rmd` vignette for more information  
  
    ```{r tidy_usada_sports, echo=FALSE}
    fs::dir_tree(
      "../inst/extdata/", 
      regexp = "sports")
    ```
    

### Data 

Below I'll import the most recent version of the tidy USADA sports data:

```{r get_recent_path-tidy-sports, eval=FALSE}
get_recent_path("../inst/extdata/", 
  type = "csv")
✔ import code pasted to clipboard!
✔ use: readr::read_csv('../inst/extdata/tidy_sports.csv')
```

```{r import-tidy_usada_sports}
tidy_usada_sports <- readr::read_csv(
  file = '../inst/extdata/tidy_sports.csv')
dplyr::glimpse(tidy_usada_sports)
```

These data have had their names standardized, and all the text has been converted to lowercase. 

## AAFs vs ADRVs

The sanctions are divided into two categories: 

- `analytic`: [Adverse Analytical Finding](https://www.usada.org/spirit-of-sport/education/alphabet-soup-results-management/#:~:text), AAF; *An AAF is a report from a WADA-accredited laboratory that identifies the presence of a prohibited substance and/or its metabolites or markers in a sample.* 

- `non-analytic`: [Non-Analytical Anti-doping Rule Violation](https://www.usada.org/spirit-of-sport/education/non-analytical-anti-doping-rule-violations/) ADRV; *a non-analytical anti-doping rule violation does not stem from a positive urine or blood sample, but instead originates from, and is substantiated by, other evidence of doping or violations by an athlete or athlete support personnel.*. 

### `substance_reason`

The `substance_reason` column contains the details of the sanction, which can include the following information: 

1. The name of the banned substance  

2. A description of the infraction (if non-analytic)  

See the examples below:

```{r steroid-view}
stringr::str_view_all(head(tidy_usada_sports$substance_reason, 25), 
  "steroid", match = TRUE)
```

```{r non-analytic-view}
stringr::str_view_all(head(tidy_usada_sports$substance_reason, 10),
  "non-analytical", match = TRUE)
```

### Sanction type

Most of the non-analytic sanctions include the terms `non-analytic`/`non-analytical`/etc., as a prefix in the `substance_reason` column.

I can pass these terms as regular expressions to create the `sanction_type` variable, which will contain two values: `non-analytical` and `analytical`

I'll save this variable in the `usada_sanction_types` dataset. 

```{r usada_sanction_types}
usada_sanction_types <- dplyr::mutate(.data = tidy_usada_sports,
    sanction_type = case_when(
      str_detect(string = substance_reason, 
        "non-analytical") ~ "non-analytical",
      !str_detect(substance_reason, 
        "non-analytical") ~ "analytical",
      TRUE ~ NA_character_
    ) 
  )
usada_sanction_types |> 
  dplyr::count(sanction_type, sort = TRUE)
```

## Analytic sanctions

Now I can filter `usada_sanction_types` to the `analytical` sanctions in `sanction_type`. 

```{r analytic_sanctions}
dplyr::filter(usada_sanction_types, 
        sanction_type == "analytical") |> nrow()
```

This is the same number of `"analytical"` values in `sanction_type`. 

## Non-analytic sanctions

```{r nonanalytic_sanctions}
dplyr::filter(usada_sanction_types, 
  sanction_type == "non-analytical") |> 
  nrow()
```

## Single vs. multipl substances 

How can I identify the single vs. multiple substances? 

Let's take a look at four different sanctions in `sanction_type_example`:

```{r sanction_type_example, echo=FALSE}
sanction_type_example <- tibble::tibble(
             sport = c("swimming","track & field",
                       "triathlon","track & field"),
  substance_reason = c("non-analytical: 3 whereabouts failures","cannabinoids",
                       "androgenic anabolic steroid; cannabinoids; anastrozole; clomiphene; gw1516",
                       "non-analytical: tampering, administration (iv), and trafficking (testosterone)"),
     sanction_type = c("non-analytical","analytical",
                       "analytical","non-analytical"))
sanction_type_example
```

We can see two analytic and two non-analytic sanctions, and each one has a single or multiple substance/reason. Fortunately, the sanctions with multiple items are separated by either semicolons (`; `), commas (`, `), or a conjunction (`and`), so I can use a regular expression to separate the items. 

```{r test-multiple_sr-sep}
dplyr::mutate(sanction_type_example, 
  multiple_sr = case_when(
    # identify the multiple_sr substances using a regular expression
    stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ TRUE,
    # negate the regular expression for the single substances
    !stringr::str_detect(substance_reason, "; |, | and | & | / ") ~FALSE,
    TRUE ~ NA)) |> 
  janitor::tabyl(multiple_sr, sanction_type)
```


The `multiple_sr` identifier can be used to separate sanctions with multiple substance/reasons from sanctions with a single substance or reason. 


### Multiple analytic substances

With the new `multiple_sr` indicator, I can separate `usada_substance_types` into single or multiple analytic substances. 

```{r usada_substance_types}
usada_substance_types <- usada_sanction_types |> 
  dplyr::mutate(multiple_sr = case_when(
    # identify the multiple_sr substances using a regular expression
    stringr::str_detect(substance_reason, "; |, | and | & | / ") ~ TRUE,
    # negate the regular expression for the single substances
    !stringr::str_detect(substance_reason, "; |, | and | & | / ") ~FALSE,
    TRUE ~ NA)) 
usada_substance_types |> 
  dplyr::count(multiple_sr, sort = TRUE)
```

How many different combinations of multiple substances are there?

```{r multiple_analytic_subs}
usada_substance_types |> 
  dplyr::filter(multiple_sr == TRUE) |> 
  dplyr::count(substance_reason, sort = TRUE, name = "Count") |> 
  count(Count, sort = TRUE)
```

Above is a 'count of counts', which tells me most of the multiple substance combinations occur only once.

## Single analytic substances

This vignette is going to focus on the sanctions with a single substance listed (we will deal with multiple substances next). 

```{r single_analytic_subs}
single_analytic_subs <- usada_substance_types |> 
  dplyr::filter(multiple_sr == FALSE) 
```

Below is a count of the single analytic substances occurring more than 5 times in the data: 

```{r tabyl-single_analytic_subs}
single_analytic_subs |> 
  janitor::tabyl(substance_reason) |> 
  janitor::adorn_pct_formatting() |> 
  dplyr::arrange(desc(n)) |> 
  dplyr::filter(n > 5)
```

## Classifying substances 

To identify the [WADA banned substances](https://www.wada-ama.org/en/prohibited-list#search-anchor), I've written `classify_wada_substances()`, a function that scans the `substance_reason` column and identifies any substances found on the [WADA list](https://www.wada-ama.org/sites/default/files/2022-09/2023list_en_final_9_september_2022.pdf). See the Appendix (bottom) for more information. 

### How it works (S1 ANABOLIC AGENTS example)

`dopingdata` stores vectors with each substance group in the WADA list (S1 substances are below): 

```{r s1_substances}
head(dopingdata::s1_substances, 10)
```

These vectors are also stored as regular expressions, which we can use to match the `substance_reason` column on:

```{r veiw-s1}
stringr::str_view_all(string = single_analytic_subs$substance_reason, 
  pattern = s1_regex, match = TRUE)
```

`classify_wada_substances()` creates a `substance_group` variable with each of the WADA classifications (these are stored in `dopingdata::wada_classes`)

```{r show-dopingdata-wada_classes}
dopingdata::wada_classes
```


### `tidy_single_aafs`

The output from `classify_wada_substances()` is stored in `tidy_single_aafs`

```{r tidy_single_aafs}
tidy_single_aafs <- classify_wada_substances(
  usada_data = single_analytic_subs, 
  subs_column = substance_reason) 
```

What `substance_group`'s appear the most?

```{r}
tidy_single_aafs |> 
  janitor::tabyl(substance_group) |> 
  janitor::adorn_pct_formatting() |> 
  dplyr::arrange(desc(n)) 
```


### S2 PEPTIDE HORMONES, GROWTH FACTORS, RELATED SUBSTANCES, AND MIMETICS

The S2 substances matches are below:

```{r veiw-s2}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s2_regex, match = TRUE)
```

### S3 BETA-2 AGONISTS

The S3 substance matches are below:


```{r veiw-s3}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s3_regex, match = TRUE)
```

### S4 HORMONE AND METABOLIC MODULATORS

The S4 substance matches are below:

```{r veiw-s4}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s4_regex, match = TRUE)
```

### S5 DIURETICS/MASKING AGENTS

The S5 substance matches are below

```{r veiw-s5}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s5_regex, match = TRUE)
```

## PROHIBITED METHODS

### M1 MANIPULATION OF BLOOD

The M1 method matches are below:

```{r veiw-m1}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = m1_regex, match = TRUE)
```

### M2 CHEMICAL AND PHYSICAL MANIPULATION

The M2 method matches are below:

```{r veiw-m2}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = m2_regex, match = TRUE)
```

### M3 GENE AND CELL DOPING

The M3 method matches are below:

```{r veiw-m3}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = m3_regex, match = TRUE)
```

## SUBSTANCES & METHODS PROHIBITED IN-COMPETITION

### S6 STIMULANTS

*All S6 STIMULANTS, including all optical isomers, e.g. `d-` and `l-` where relevant, are prohibited.*

The S6 substance matches are below:

```{r veiw-s6}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s6_regex, match = TRUE)
```

### S7 NARCOTICS

The S7 substance matches are below:

```{r veiw-s7}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s7_regex, match = TRUE)
```

### S8 CANNABINOIDS

*The following cannabinoids are prohibited: Natural cannabinoids, e.g. cannabis, hashish and marijuana; Synthetic cannabinoids e.g. Δ9-tetrahydrocannabinol (THC) and other cannabimimetics.*

The S8 substance matches are below:

```{r veiw-s8}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s8_regex, match = TRUE)
```

### S9 GLUCOCORTICOIDS

The S9 substance matches are below:

```{r veiw-s9}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = s9_regex, match = TRUE)
```

### P1 BETA BLOCKERS (prohibited In-Competition only)

*Beta-blockers are prohibited In-Competition only, in the following sports, and also prohibited Out-of-Competition where indicated.*

+ archery (wa)
+ automobile (fia)
+ billiards (all disciplines) (wcbs)
+ darts (wdf)
+ golf (igf)
+ shooting (issf, ipc)
+ skiing/snowboarding (fis) in ski jumping, freestyle aerials/halfpipe and snowboard halfpipe/big air
+ underwater sports (cmas) in constant-weight apnoea with or without fins, dynamic apnoea with and without fins, free immersion apnoea, jump blue apnoea, spearfishing, static apnoea, target shooting, and variable weight apnoea

The S8 substance matches are below:

```{r veiw-p1}
stringr::str_view_all(string = tidy_single_aafs$substance_reason, 
  pattern = p1_regex, match = TRUE)
```

## Unclassified substances 

The substances that were not classified using the standard WADA list will be stored in `unclass_single_aaf_substances`.

```{r unclass_single_aaf_substances}
unclass_single_aaf_substances <- tidy_single_aafs |> 
  filter(is.na(substance_group)) 
unclass_single_aaf_substances |> 
  count(substance_reason, sort = TRUE)
```

I can use some additional regular expressions with the `classify_substance()` function.

### `"elevated t/e"` & `3 whereabouts failures`

```{r classify_substance-unclassified}
classify_substance(
  df = unclass_single_aaf_substances, 
  subs_col = substance_reason, 
  subs = "elevated t/e", 
  var = "analytic",
  val = "S1 ANABOLIC AGENTS",
  wb = FALSE) |> 
  classify_substance(
  subs_col = substance_reason, 
  subs = "3 whereabouts failures", 
  var = "non_analytic",
  val = "3 whereabouts failures",
  wb = FALSE) |> 
  dplyr::filter(!is.na(analytic) | !is.na(non_analytic)) |> 
  dplyr::select(athlete, sanction_date, substance_reason, analytic, non_analytic)
```

I can use these to recode the original values--start with the `3 whereabouts failures`

`yair rodriguez` should be non-analytical, so we can create a tiny tibble to bind back to `usada_substance_types` (after removing `rodriguez, yair`)

```{r miss_classified_sanction_type}
# get rodriguez, yair (3 whereabouts failures)
miss_classified_sanction_type <- tidy_single_aafs |> 
  filter(athlete == "rodriguez, yair" & 
            sanction_date == lubridate::as_date("2020-12-03")) |> 
  dplyr::mutate(sanction_type = if_else(sanction_type == "analytical", 
    true = "non-analytical", sanction_type)) |> 
  dplyr::select(-substance_group)
usada_substance_types <- usada_substance_types |> 
  dplyr::filter(athlete != "rodriguez, yair" & 
            sanction_date != lubridate::as_date("2020-12-03")) |> 
  dplyr::bind_rows(miss_classified_sanction_type)
usada_substance_types |> 
  dplyr::filter(athlete == "rodriguez, yair") |> 
  dplyr::select(athlete, sanction_date, substance_reason, sanction_type)
```

Now we can address the `elevated t/e` (I classify this as `"S1 ANABOLIC AGENTS"`)

```{r miss_classified_sanction_s1}
tidy_single_aafs <- tidy_single_aafs |> 
  dplyr::mutate(
    substance_group = case_when(
      stringr::str_detect(substance_reason, "elevated t/e") ~ "S1 ANABOLIC AGENTS")) 

tidy_single_aafs |> 
  dplyr::filter(str_detect(athlete, "kirk")) |> 
  dplyr::select(athlete, sanction_date, substance_reason, sanction_type)
```

## Export sanction types

Export analytic sanctions (`usada_sanction_types`):

```{r export_extdata-usada_sanction_types}
export_extdata(
  x = usada_sanction_types, 
  inst_path = "../inst/", 
  raw = FALSE
)
```


```{r vrfy-usada_sanction_types}
fs::dir_tree('../inst/extdata/', regexp = "types")
```


## Export substance types

Export `usada_substance_types`

```{r export_extdata-usada_substance_types}
export_extdata(
  x = usada_substance_types, 
  inst_path = "../inst/", 
  raw = FALSE
)
```

```{r vrfy-usada_substance_types}
fs::dir_tree('../inst/extdata/', regexp = "usada_substance_types")
```


## Export tidy classified (single) substances 

Export classified single substances (`tidy_single_aafs`):

```{r export_extdata-tidy_single_aafs}
export_extdata(
  x = tidy_single_aafs, 
  inst_path = "../inst/", 
  raw = FALSE
)
```

```{r vrfy-tidy_single_aafs}
fs::dir_tree("../inst/extdata/", regexp = "tidy_single_aafs")
```

# Appendix 

## WADA list 

A copy of the original WADA substances is available in `dopingdata::WADA`

```{r WADA, echo=FALSE}
dopingdata::WADA |> 
  dplyr::count(groups, type, 
    name = "Substance Count", 
    sort = TRUE) |> 
  ungroup() |> 
  knitr::kable(
    caption = "counts of substances per group in dopingdata::WADA") |> 
  kableExtra::kable_minimal()
```
